Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-30 paper.tex

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.
Two technological developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing ---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2015scrm}
were necessary to simulate whole chromosomes.

The second development that has made the majority of
present day simulators inapplicable to modern data sets is the
exponential increase in sample size. Human datasets such
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2019variation} now contain hundreds of
thousands of genomes, and there is every reason to believe that
researchers will be soon be routinely working with
datasets containing millions of samples~\citep{stephens2015big}.
Classical simulators such as \ms\ and even fast approximate simulators
such as \scrm\ simply cannot cope with this scale. Even if we
spend months of CPU time on running simulations with millions of samples,
the output would require terabytes of storage space to store and
days of CPU time to parse~\citep{kelleher2016efficient}.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations.
Through the use of efficient data structures, it is
now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} or
VCF~\citep{danecek2011variant} based formats.
This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

% JK: Earlier notes on this para
% - Msprime 1.0 combines the features of many different simulators and is
%   far more efficient than all of them. The combination of a Python API,
%   an extremely efficient storage format, and a well-engineered and
%   extensible codebase changes the game. Stdpopsim builds on this base
%   to provide easy access to standard models of population genetics.
%

>>>>

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating the samples to each other.
The coalescent has proved to be a powerful model,
and it is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.
Two technological developments of recent years, however,
pose major challenges to the majority of these methods.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2015scrm}
were necessary to simulate whole chromosomes.
The second development reducing the utility of many simulators is the
exponential increase in sample size. Human datasets such
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm\ simply cannot cope with such a large number of samples.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.
